\section{Distribution Extraction} 
\label{sec:distex}


Given a program with a \passert $e$ and either a concrete input or a
distribution over inputs,
\tool performs a probabilistic evaluation
by building and optimizing a Bayesian-network representation of the statements
required to evaluate the \passert.  This section describes
distribution extraction, which is the first step in this process.
Distribution extraction produces a symbolic Bayesian network representation
that corresponds to the slice of the program contributing to $e$.
\tool treats randomness as symbolic and
deterministic components as concrete.
The process is similar to symbolic execution and to lazy evaluation in
functional languages.

\paragraph{Distributions as Symbolic Values}
\mayhap performs a forward pass over the program, concretely
evaluating deterministic computations and introducing symbolic
values---probability-distribution expression trees---to represent
probabilistic values. For example, the following statement: 
%
\begin{lstlisting}[language=Python]
  a = b + 2
\end{lstlisting}
%
computes $a$ concretely when $b$ is not probabilistic.  If, prior to the above
statement, the program assigns \lil{b = 5}, then we perform the
addition and set $a=7$.  However, if \lil{b = gaussian()}, we
add a node to the Bayesian network, representing $b$
symbolically as a Gaussian distribution.  We then create a sum node for
$a$ with two parents: $b$'s Gaussian  and $2$'s constant (point mass) distribution.

As this mixed symbolic and concrete execution proceeds, it eagerly
evaluates any purely deterministic statements but builds a Bayesian-network
representation
of the forward slice of any probabilistic statements.  This
process embodies a symbolic execution in which the symbolic values are
probability distributions. Our approach differs from typical symbolic
execution
in how it handles control flow (see below).  

When the analysis reaches a statement \passert $e$, \mayhap records the Bayesian network
rooted at $e$. It then optimizes the network and samples the
resulting distribution.
Compared to sampling the entire program repeatedly, sampling the extracted
distribution can be more efficient even without optimizations since it
eliminates redundant, non-probabilistic
computation. 

%  the work
% that would be wasted in a straightforward sampling approach to
% evaluating \passerts---by sampling the specialized distribution (as in
% Section~\ref{sec:sample}), we sample only the portion of the program
% that changes from execution to execution.

\label{sec:conditionals}


\paragraph{Conditionals}
% Any symbolic execution technique must address the possibility of control flow
% based on symbolic values. 
When conditionals and loops are based on purely
concrete values, distribution extraction proceeds down one side
of the control flow branch as usual. When conditions operate on probabilistic variables, the analysis
must capture the effect of both branch directions.

To analyze the probability distribution of a conditional statement,
we produce conditional probabilities
based on control-flow divergence. For example, consider this simple
program:
%
\begin{lstlisting}[language=Python]
  if a:  b = c   else:  b = d
\end{lstlisting}
%
in which $a$ is probabilistic. Even if both $c$
and $d$ are discrete, the value of $b$ is probabilistic since it depends on
the value of $a$. We can write the conditional probability distributions $\prob{B}$
for $b$ conditioned on both possible values for $a$:
%
\begin{align*}
  \Pr[B \;|\; A = \mathrm{true}] &= \Pr[C]\\
  \Pr[B \;|\; A = \mathrm{false}] &= \Pr[D]
\end{align*}
%
Instead, to simplify the representation of probabilities and to enable more
straightforward analysis, we \emph{marginalize} the condition $a$ to
produce an unconditional distribution for $b$. Using marginalization, we
write the unconditional distribution $\prob{B}$ as:
%
\begin{align*}
    \Pr[B] &= \sum_a \Pr[B\;|\; A = a] \prob{A = a} \\
           &= \Pr[B \;|\; A = \mathrm{true}] \prob{A = \mathrm{true}} \\
           & \qquad {}
            + \Pr[B \;|\; A = \mathrm{false}] \prob{A = \mathrm{false}} \\
           &= \Pr[C] \cdot \prob{A=\mathrm{true}} \\
           & \qquad {}
            + \Pr[D] \cdot (1 - \prob{A=\mathrm{true}})
\end{align*}
%
This expression computes the distribution for $b$ as a function of the
distributions for $a$, $c$, and $d$. Intuitively, the probabilistic evaluation rewrites
the condition to read \lil{b = a * c + (1 - a) * d}. This algebraic
representation enables some optimizations described
in Section~\ref{sec:expectation}.

\label{sec:loops}


\paragraph{Loops and External Code}
Loops with probabilistic conditions can, in general, run for an
unbounded number of iterations. Representing unbounded execution would
induce cycles in our graphical
model and violate the acyclic definition of a Bayesian network. For
example, consider a loop that accumulates samples and exits when the
sum reaches a threshold:
%
\begin{lstlisting}[language=Python]
  v = 0.0
  while v < 10.0:
    v += random.uniform(-0.5, 1.0)
\end{lstlisting}
%
If the random sample is negative in every iteration, then the
loop will never exit. The probability of this divergence is small but non-zero.

Prior work has dealt with probabilistic loops by restricting the program
to linear operators~\cite{sriram-pldi}. \tool relaxes this
assumption by treating a loop as a
black box that generates samples (i.e., the loop may run for an
unbounded but finite number of iterations), similar to a known
probability distribution such as \code{random.uniform}.
This representation avoids creating cycles.
In particular, \tool represents a loop body with a
\emph{summary node}, where variables read by the loop are
edges \emph{into} the node and variables written by the loop are edges \emph{out} of the node.

In practice, many loops in probabilistic programs have
non-probabilistic bounds. For example, we evaluated an image filter
(\bench{sobel}) that loops over the pixel array and applies a probabilistic
convolution to each window. The nested loops resemble:
%
\begin{lstlisting}[language=Python]
  for x in 0..width:
    for y in 0..height:
      filter(image[x][y])
\end{lstlisting}
%
While the computed pixel array contains probabilistic data, the dimensions
\code{width} and \code{height} are fixed
for a particular image. \tool extracts complete distributions from these common
concrete-bounded loops without black-box sampling.

\tool uses a similar black-box mechanism when interfacing with library code whose
implementation is not available for analysis---for example, when passing a
probabilistic value to the \code{cos()} function from the C standard library.
This straightforward approach prevents statistical optimizations inside the
library function or loop body but lets \tool analyze more programs.

\paragraph{Analyzing Loops with Probabilistic Path Pruning}
We propose another way to analyze loops with probabilistic bounds by
building on the path pruning techniques used in traditional symbolic
execution.  Typically, path pruning works by proving that some paths
are infeasible. If the analysis determines that a path constraint
is unsatisfiable, it halts exploration of that path.  Probabilistic evaluation
instead needs to discover when a given path is \emph{unlikely} rather
than impossible, i.e., when the conditions that lead to following this
path at run time have a probability that falls below a
threshold.  We propose tracking a path probability expression for each
explored path and periodically sampling these distributions to
prune unlikely paths.
This extension handles general probabilistic
control flow in programs that are likely to terminate eventually.
Intuitively, the more iterations the loop
executes, the less likely it is to execute another iteration.
Programs with a significant probability of running forever before
reaching a \passert can still prevent the analysis from terminating,
but this behavior likely indicates a bug.
We leave the evaluation of
this more precise analysis to future work.

% Assume a simple probabilistic program with a loop that lacks
% concrete bounds (i.e., the bounds are statistical).

% x = uniform(0,1)
% w = 0
% Label:
%   w += x + 1
%   if w < 100 goto Label
% z = w * 10
% passert z < 10

% Note we cannot naively represent this program as a Bayesian
% network because the loop induces a cycle in the induced graphical
% model.

% However, if we wrap the body of the loop into its own super-node
% we can remove the cycle in the graphical model and turn it into a
% Bayesian network---variables read within the body of the loop
% will be edges \emph{into} the super-node and variables written by
% the body of the loop will be edges \emph{out} of the super-node.
% This provides a mechanism to handle loops.  Of course, \tool is
% now unable to optimize the contents of the super-node and must
% sample from it but as we demonstrate in Section~\cite{sec:eval}
% sampling is still efficient.

% To see this in practice, consider the prior code. We represent
% the loop's basic block, denoted by Label, as a super-node with
% edges from the random variable X into the super-node and edges
% from the super-node into Z.

% The passert asks whether Z < 10 and so \tool verifies the passert
% as true if E[Z < 10] > 0.5 and false if E[Z < 10] < 0.5.  In
% particular, we use (i) exact methods which use probabilistic laws
% to prove E[X<10] < 0.5 or (ii) we use hypothesis testing to
% estimate E[Z < 10] and bound error in that estimate.
