I acknowledge that pursuing a Ph.D.~is only made worthwhile by the people who
surround you along the way.

This dissertation describes work done by a staggering array of amazing
researchers who are not me.
My co-authors are too numerous to list exhaustively here, but they can be
found hiding in the references~\cite{enerj, decaf, passert, approxstorage}.
Notably, Werner Dietl masterminded EnerJ's semantics,
Karin Strauss is the memory technology expert, and
Pavel Panchekha is behind the formalism for probabilistic assertions.
DECAF in Chapter~\ref{ch:decaf} is Brett Boston's research
project for the honors designation on his bachelor's degree.
He did all the hard work; I played the role of a meddling mentor.

Most of all, I have had the unique honor of being advised by the two best
mentors in the computer-science universe, Luis Ceze and Dan Grossman.
Luis is an unstemmable font of creativity, a fierce advocate, and a true
research visionary.
Dan is a tireless champion of good taste, an enemy of bullshit, and a stalwart
encyclopedia of great advice.
No page in this disseration could exist without either of them.

Thank you to my staggeringly brilliant collaborators.
Emily Fortuna lent a hand early on, along with Danushen Gnanapragasam.
Hadi Esmaeilzadeh took the first plunge into approximate hardware with us and
his advisor, Doug Burger.
Jacob Nelson, Thierry Moreau, Andre Baixo, Ben Ransford, Mark Wyse, and
Michael Ringenburg were all instrumental to defining the approximaiton agenda.
Ben Wood inducted me into research in the best way possible.

Thank you to the UW undergraduates who withstood my mentoring:
Luyi Liu,
Chengfeng Shi,
Joshua Yip,
Brett Boston,
Wenjie (Marissa) He,
Finn Parnell, and
Danushen Gnanapragasam.
Their talents are unbounded.

Thank you to my research groups, Sampa and PLSE.\ Extra-large thanks to
my role models in Sampa: Joe Devietti, Brandon Lucia and Jacob Nelson.
They defined computer architecture for me.
A very real thank you to Sampa's own
Owen Anderson,
Katelin Bailey,
Tom Bergan,
James Bornholt,
Hadi Esmaeilzadeh,
Emily Fortuna,
Brandon Holt,
Nick Hunt,
Vincent Lee,
Eric Mackay,
Thierry Moreau,
Brandon Myers,
Ben Ransford,
Michael Ringenburg,
Ben Wood,
and
Carlo del Mundo.
They endured my many terrible ideas and made grad school bearable.
Thank you to Melody Kadenko, without whom everything would immediately
collapse.

Thank you to my menagerie of mentors:
Mark Oskin, the godfather of the Sampa group,
Călin Caşcaval from Qualcomm,
and
Karin Strauss, Todd Mytkowicz, and Kathryn McKinley from Microsoft Research.
Thank you to my patient PhD committee, including
Maryam Fazel, Eric Klavins, Vivesh Sathe, and Hank Levy, who deserves a whole
second category of gratitude.
Thank you to the entire faculty at UW CSE, who said so many encouraging things
over the years, and who occasionally endured merciless holiday-party
lampooning.

Many exceptionally emphatic thanks to Lindsay Michimoto, who did so many
things to make grad school work for me that they cannot possibly fit here, and
whose constant kindness and inexhaustible work ethic continue to inspire me.
Thank you to my surprisingly helpful officemate-mentors, Ivan Beschastnikh and
Chloé Kiddon.

Thank you to my advisors and mentors at Harvey Mudd,
Ran Libeskind-Hadas,
Melissa O'Neill,
Bob Keller,
and Geoff Kuenning,
who are more responsible than anyone for getting me into this in the first
place.
Double thanks to Ran, from whom I first really learned what computer science
is, and who is still an inspiration every day.
Triple thanks to Ran for his prescient advice to pursue anything else over
theory.

Thank you to my parents, Elizabeth Dequine and Ed Sampson~III, for everything
and also for their infinite tolerance of my inexplicable interests.
Infinite and everlasting thanks to the magnificent Ariana Taylor-Stanley, the
basis of the whole thing.
